---
title: "final_project_code"
author: "Angelina A., Heonmi K., Caroline S., and Zilin W."
date: "Spring 2026"
output:
  pdf_document: default
  html_document: default
---

```{r, include = F}
library("readxl") 
# May need to install.packages("readxl")
```


## 1. Reading in the data

```{r}
read_data <- read_excel("./stat571B_reading_data.xlsx")

read_data['total_time'] = read_data['Time (minutes)']*60 + read_data['Time (seconds)']

read_data['format'] = c("Laptop", "Laptop", "Print", 
                        "Laptop", "Print", "Print",
                        "Print", "Print", "Laptop", 
                        "Print", "Laptop", "Laptop",
                        "Laptop", "Print", "Print", 
                        "Laptop", "Print", "Laptop",
                        "Print", "Laptop", "Laptop",
                        "Print", "Laptop", "Print",
                        "Laptop", "Print", "Laptop",
                        "Laptop", "Laptop", "Print",
                        "Print", "Laptop", "Print",
                        "Print", "Print", "Laptop")

read_data["music"] = c("no", "yes", "no", 
                       "yes", "no", "yes", 
                       "no", "yes", "no", 
                       "yes", "no", "yes", 
                       "no", "no", "yes",
                       "yes", "yes", "no", 
                       "no", "no", "yes", 
                       "yes", "yes", "no", 
                       "no", "yes", "yes",
                       "yes", "no", "no",
                       "no", "yes", "yes",
                       "yes", "no", "no")
                
read_data
```


## 2. Analysis

We used ANOVA with participant as a blocking factor to assess the significance of the experimental factors, as required. Additionally, linear mixed-effects models were fitted with participant as a random effect to account for within-subject correlation.

We fit factorial models with format, music, and their interaction as primary fixed effects. Because participants each read multiple passages, participant was included to account for within-subject variability. Since only three text passages were used, Text was included as a fixed effect rather than a random effect to adjust for possible passage-specific differences.

```{r include=FALSE}
# format
read_data$format <- factor(read_data$format)
read_data$Text <- factor(read_data$Text)
read_data$music  <- factor(read_data$music)
read_data$Participant <- factor(read_data$Participant)

# library
library(car)
library(lme4)
library(lmerTest)  
```

### Model 1: Time

```{r}
# ANOVA for Reading Time
model_time <- lm(total_time ~ format * music + Text + Participant + Participant:music, 
                 data = read_data)
Anova(model_time, type = "III")

## Mixed Effects Model 
model_time_lmer <- lmer(
  total_time ~ format * music +
    (1 | Text) +
    (1 + music | Participant),
  data = read_data
)
anova(model_time_lmer)
```



### Model 2: Comprehension

```{r}
# ANOVA for Comprehension
model_comp <- lm(`Comprehension score (%)` ~ format * music + Text + 
                   Participant + Participant:music, 
                  data = read_data)
Anova(model_comp, type = "III")

## Mixed Effects Model 
model_comp_lmer <- lmer(
  `Comprehension score (%)` ~ format * music +
    (1 | Text) +
    (1 + music | Participant),
  data = read_data
)
anova(model_comp_lmer)
```



### Diagnostics for ANOVA

```{r}
par(mfrow = c(2,2))
plot(model_time)

par(mfrow = c(2,2))
plot(model_comp)
```

