---
title: "final_project_code"
author: "Angelina A., Heonmi K., Caroline S., and Zilin W."
date: "Spring 2026"
output:
  pdf_document: default
  html_document: default
---

```{r, include = F}
library("readxl") 
# May need to install.packages("readxl")
```


## 1. Reading in the data

```{r}
read_data <- read_excel("./stat571B_reading_data.xlsx")

read_data['total_time'] = read_data['Time (minutes)']*60 + read_data['Time (seconds)']

read_data['format'] = c("Laptop", "Laptop", "Print", 
                        "Laptop", "Print", "Print",
                        "Print", "Print", "Laptop", 
                        "Print", "Laptop", "Laptop",
                        "Laptop", "Print", "Print", 
                        "Laptop", "Print", "Laptop",
                        "Print", "Laptop", "Laptop",
                        "Print", "Laptop", "Print",
                        "Laptop", "Print", "Laptop",
                        "Laptop", "Laptop", "Print",
                        "Print", "Laptop", "Print",
                        "Print", "Print", "Laptop")

read_data["music"] = c("no", "yes", "no", 
                       "yes", "no", "yes", 
                       "no", "yes", "no", 
                       "yes", "no", "yes", 
                       "no", "no", "yes",
                       "yes", "yes", "no", 
                       "no", "no", "yes", 
                       "yes", "yes", "no", 
                       "no", "yes", "yes",
                       "yes", "no", "no",
                       "no", "yes", "yes",
                       "yes", "no", "no")
                
read_data
```


## 2. Analysis

We used ANOVA with participant as a blocking factor to assess the significance of the experimental factors, as required. Additionally, linear mixed-effects models were fitted with participant as a random effect to account for within-subject correlation.

We fit factorial models with format, music, and their interaction as primary fixed effects. Because participants each read multiple passages, participant was included to account for within-subject variability. Since only three text passages were used, Text was included as a fixed effect rather than a random effect to adjust for possible passage-specific differences.

```{r include=FALSE}
# format
read_data$format <- factor(read_data$format)
read_data$Text <- factor(read_data$Text)
read_data$music  <- factor(read_data$music)
read_data$Participant <- factor(read_data$Participant)

# library
library(car)
library(lme4)
library(lmerTest)  
```

### Model 1: Time

```{r}
# ANOVA for Reading Time
model_time <- lm(total_time ~ format * music + Text + Participant, data = read_data)
Anova(model_time, type = "III")

## Mixed Effects Model (Supplementary Analysis)
lmer_time <- lmer(total_time ~ format * music + Text + (1 | Participant), data = read_data)
anova(lmer_time)
```

For reading time, neither format nor the interaction between format and music was statistically significant. Music showed a marginal effect in the ANOVA model and was statistically significant in the mixed-effects model, suggesting a potential influence of music on reading time after accounting for within-subject variability.
The effect of Text was not statistically significant, indicating no strong evidence of differences in reading time across the three passages.

### Model 2: Comprehension

```{r}
# ANOVA for Comprehension
model_comp <- lm(`Comprehension score (%)` ~ format * music + Text + Participant, data = read_data)
Anova(model_comp, type = "III")

## Mixed Effects Model (Supplementary Analysis)
lmer_comp <- lmer(`Comprehension score (%)` ~ format * music + Text + (1 | Participant), data = read_data)
anova(lmer_comp)
```

For comprehension, the interaction between format and music was statistically significant in the ANOVA model and marginally significant in the mixed-effects model. This indicates that the effect of format on comprehension depends on the presence of music, and therefore the main effects of format and music should not be interpreted independently.
The main effects of format and music were not statistically significant on their own. The effect of Text was also not statistically significant, suggesting no strong evidence of differences in comprehension across passages.

### Diagnostics for ANOVA

```{r}
par(mfrow = c(2,2))
plot(model_time)

par(mfrow = c(2,2))
plot(model_comp)
```

Diagnostic plots for both models indicate no major violations of model assumptions.